% Dorian's notes and comments
% Personal scratch file --- not included in main.tex

\section*{Notes: Building a Simple Snake Game in Python}

This note explains a small Snake game written in Python using the
\texttt{pygame} library. The source file is \texttt{snake.py} and lives
outside the Overleaf repository (in the parent project folder).

\subsection*{What you need to run it}

Python 3 and the \texttt{pygame} library. Install pygame once with:

\begin{quote}
\texttt{pip install pygame}
\end{quote}

Then launch the game with \texttt{python snake.py}. Steer the snake with
the arrow keys or W/A/S/D. Eat the red squares to grow. Don't run into
the walls or yourself. Press \texttt{R} after a game over to restart.

\subsection*{The core idea: a grid of cells}

Instead of moving the snake pixel by pixel, the playing field is treated
as a grid of cells. Each cell is 20 pixels wide. The grid is 30 cells
across and 20 cells tall, which gives a window of $600 \times 400$
pixels. Everything in the game lives on this grid: the snake is a list
of cell coordinates, and the food is a single cell coordinate. Thinking
in grid units rather than pixels makes collision checks and movement
simple integer comparisons.

\subsection*{Representing the snake}

The snake is stored as a Python list of \texttt{(x, y)} tuples. The
first element is the head, the rest are the body, the last is the tail.
A direction is also stored as a tuple, for example \texttt{(1, 0)} means
``one cell to the right per step'' and \texttt{(0, -1)} means ``one cell
up per step''.

\subsection*{The game loop}

The whole game runs inside one big \texttt{while True} loop. Each pass
through the loop is one ``tick'' of the game and does four things in
order:

\begin{enumerate}
    \item \textbf{Handle input.} Read all keyboard events since the last
    tick. If the user pressed an arrow key, remember the new direction
    in a separate variable called \texttt{pending}. The active direction
    is only updated at the start of the next tick. This avoids a classic
    bug where pressing two keys quickly (for example right then down
    while moving right) could cause the snake to reverse into itself.
    A guard also rejects any key that would directly reverse the current
    direction.

    \item \textbf{Move the snake.} Compute a new head position by adding
    the direction vector to the current head. Then check for collisions
    (described below). If the move is legal, insert the new head at the
    front of the list. If the new head landed on the food, increase the
    score and place new food; otherwise, remove the tail. Adding a head
    and removing a tail makes the snake appear to slide forward one
    cell without rebuilding the whole list.

    \item \textbf{Draw.} Fill the screen black, then draw every snake
    segment as a green square and the food as a red square. The head is
    drawn slightly brighter than the body so you can see which way the
    snake is facing. The score is drawn in the top-left corner. If the
    game is over, a centred message is drawn over the field.

    \item \textbf{Pace the loop.} Call \texttt{clock.tick(FPS)} with
    \texttt{FPS = 10}, which forces the loop to run roughly ten times
    per second. Without this the loop would run as fast as the CPU
    allows and the snake would be uncontrollable.
\end{enumerate}

\subsection*{Collisions}

Two kinds of collisions end the game:

\begin{itemize}
    \item \textbf{Wall:} the new head coordinate falls outside the grid,
    that is its $x$ is not in $[0, 30)$ or its $y$ is not in $[0, 20)$.
    \item \textbf{Self:} the new head coordinate appears anywhere else
    in the snake's list of segments.
\end{itemize}

If either happens, a \texttt{game\_over} flag is set. The drawing code
keeps running so the final frame stays visible, and a key handler waits
for \texttt{R} to restart.

\subsection*{Placing food}

Food is placed by picking a random cell and checking it is not already
occupied by the snake. If it is, another random cell is picked. For a
small grid this rejection-sampling approach is fast enough and avoids
having to track every free cell separately.

\subsection*{Why this design}

Snake is a good first game because every interesting concept stays
simple:

\begin{itemize}
    \item State is just a list of tuples, an integer score, and a
    direction vector --- no classes or inheritance needed.
    \item Movement is one vector addition per tick.
    \item Collision is one bounds check and one membership test.
    \item The render step is a handful of coloured rectangles.
\end{itemize}

The whole game fits comfortably in around a hundred lines and is a
useful template for any other grid-based game (Tetris, Pac-Man, etc.):
keep a model of the world as plain data, advance it one tick at a time,
and redraw from scratch each frame.

\section*{Notes: A Mini Galaxy Game}

This chapter builds directly on the snake game. Same tools
(\texttt{pygame}), same overall structure (one big game loop, plain
data for state, redraw each frame), but the world is now continuous
instead of grid-based, the character can jump, gravity behaves
strangely, and there are enemies with their own behaviour. The source
file is \texttt{galaxy.py}, again in the parent project folder beside
\texttt{snake.py}.

\subsection*{A note on scope}

Super Mario Galaxy is a full 3D platformer with animated characters,
fluid camera work and complex physics --- well beyond a single Python
file. What this game keeps from Galaxy is its \emph{signature
mechanic}: you walk around the outside of a small planet, and gravity
pulls toward the planet's centre rather than ``down''. Everything else
(camera, art, level design) is intentionally minimal. The result is a
small 2D homage that captures the idea, not a clone.

\subsection*{The key new idea: polar coordinates}

In the snake game, a position was a grid cell \texttt{(x, y)}. Here a
position is described relative to the planet's centre using two
numbers:

\begin{itemize}
    \item $\theta$ (theta) --- the angle around the planet, in radians.
    Walking around the surface means changing $\theta$.
    \item $r$ --- the distance from the planet's centre. When the
    character stands on the ground, $r$ equals the planet's radius
    plus the character's size. Jumping makes $r$ larger; gravity pulls
    it back down.
\end{itemize}

Storing positions this way makes the physics very natural. ``Walking''
is just changing $\theta$. ``Falling toward the planet'' is just $r$
decreasing. ``Standing on the ground'' is just $r$ being clamped to a
fixed value.

When it is time to draw, we convert back to screen pixels with the
standard polar-to-Cartesian formulas:
\[
x = c_x + r \cos\theta, \qquad y = c_y + r \sin\theta,
\]
where $(c_x, c_y)$ is the centre of the planet on the screen.

\subsection*{Radial gravity}

Each tick, the radial velocity $v_r$ has gravity subtracted from it:
$v_r \leftarrow v_r - g$. Because $v_r$ is along the line from the
planet's centre outward, subtracting from it is always a pull
\emph{inward}. The player therefore falls toward the planet from
whichever side they are on --- the top, the bottom, or anywhere
in between. Then $r \leftarrow r + v_r$. If $r$ drops below
ground level we clamp $r$ to the ground and set $v_r = 0$; the player
is now standing.

\subsection*{Jumping and air control}

A jump is one line: when the player presses space and is on the
ground, set $v_r$ to a positive value (the jump strength). Gravity
takes care of the rest --- $v_r$ shrinks each tick until the player
hits the ground again. While in the air the player keeps the
horizontal momentum they had on the ground, with a small amount of
``air control'' allowed (steering mid-jump), capped so the player can
not accelerate beyond walking speed.

\subsection*{Walking around the planet}

When the player is on the ground, pressing left/right sets the
angular velocity $v_\theta$ directly: $v_\theta = \text{input} \times
\text{walkSpeed}$. With a fixed camera the same key always rotates
around the planet in the same direction, so the player walks all the
way around the world without ever changing how the controls work.

A small visual detail makes this readable: the character is drawn as a
red circle with a smaller darker circle stuck to the side facing away
from the planet. That smaller circle acts as a ``hat'', and because it
is always offset along $(\cos\theta, \sin\theta)$ it automatically
points away from the planet --- so the character visibly stands
upright whether they are on top of the planet, the side, or upside
down on the bottom.

\subsection*{Stars (the objective)}

Five stars are placed at random angles around the planet, each at a
slightly different height above the surface. They are drawn as yellow
circles that pulse softly (their radius is modulated by a sine wave on
the tick counter). A star is collected when the screen-space distance
between it and the player drops below the sum of their radii. Picking
up the last star sets a \texttt{won} flag and the game displays a win
screen.

\subsection*{Enemies and stomping}

Three enemies patrol the surface. Each has an angle $\theta$ and a
direction ($+1$ or $-1$); each tick the angle increases by direction
$\times$ enemy speed. Collisions are checked exactly like with the
stars --- by screen-space distance. What happens on a collision
depends on the player's vertical state:

\begin{itemize}
    \item If the player is meaningfully higher than the enemy (i.e.
    further from the planet centre) and currently moving toward the
    planet (i.e.\ $v_r < 0$, falling), it counts as a stomp: the
    enemy dies, and the player bounces upward by setting $v_r$ to a
    positive value.
    \item Otherwise it counts as a hit. The player loses one life and
    is bounced away. Three hits ends the game.
\end{itemize}

This is the same rule as in many platformers: ``jumping on top of an
enemy is good; running into one from the side is bad''. Encoded in
polar coordinates it is just a comparison of $r$ and a sign check on
$v_r$.

\subsection*{Background starfield}

To make the scene feel more like space, a few hundred fixed pixels are
scattered across the screen at startup, painted as tiny white dots
behind the planet. This is purely decorative --- they do not move,
collide, or interact --- but it changes how the game feels for almost
no extra code.

\subsection*{What this adds on top of snake}

Comparing the two games is a tidy way to see what changes when a game
gets a little more ambitious:

\begin{itemize}
    \item \textbf{Positions} go from a single grid cell to two real
    numbers $(\theta, r)$.
    \item \textbf{Movement} goes from a fixed-direction step every
    tick to continuous velocities that change under gravity.
    \item \textbf{Collisions} go from list membership to distance
    checks between circles.
    \item \textbf{Objects} multiply: a player, several enemies with
    their own state, several stars, a planet, a background. Each
    becomes either a small dictionary or a list of dictionaries.
\end{itemize}

The overall recipe, though, is unchanged from snake: keep the world
as plain data, advance it one tick at a time inside a single loop,
and redraw the whole screen each frame. Any future games --- a longer
platformer, a top-down shooter, a small RPG --- can grow from exactly
this template.

\section*{Notes: An Explorable Mini Galaxy World}

The previous chapter had one planet. This chapter extends it to a
small \emph{world} you can traverse: a large starting planet, a
smaller target planet, three floating mini-platforms in between, a
glowing launch star that flings you across open space, and a purple
warp tube that takes you back. The source file is \texttt{galaxy2.py},
again in the parent project folder.

\subsection*{A note on scope, again}

Mario Galaxy's traversal is gorgeous: launch stars warp Mario between
planets, pull stars yank him along curved paths, pipes drop him onto
new planetoids, gravity flips as the camera rotates. Replicating any
of that in 3D is out of session scope. What this chapter keeps is the
\emph{feeling} of those mechanics in flat 2D: multiple worlds with
their own gravity, scripted impulses that move you between them, and
warp pipes that connect distant points. Everything else is sketched
in simple coloured circles.

\subsection*{From one body to many}

The Mini Galaxy game stored the player in polar coordinates
$(\theta, r)$ centred on \emph{the} planet --- because there was only
one. As soon as there are several round bodies of different sizes and
positions, a single $(\theta, r)$ no longer makes sense. The new game
stores the player in plain Cartesian coordinates $(x, y)$, and the
``which body am I on'' question is answered by a separate field
called \texttt{ground\_body} that names which planet or platform the
player is currently standing on. While the player is on the ground,
the game still tracks an angle $\theta$ for that body, so walking
around a planet still works exactly as it did before --- just
attached to whichever body \texttt{ground\_body} points to.

\subsection*{Walking, jumping, falling}

These three states stay structurally the same as in the previous
chapter, just made body-agnostic:

\begin{itemize}
    \item \textbf{On the ground} on body $B$: arrow keys change
    $\theta$ relative to $B$, and the player's $(x, y)$ is recomputed
    each tick as $B.\text{centre} + (B.r + \text{playerSize}) \cdot
    (\cos\theta, \sin\theta)$. The player's velocity is zero.
    \item \textbf{Jumping}: the player's velocity becomes a vector
    pointing radially outward from $B$, scaled by the jump speed. The
    state flips to ``in air'' and $B$ is forgotten until the player
    lands somewhere.
    \item \textbf{In air}: gravity (described next) accelerates the
    velocity, position advances by the velocity, and a landing check
    looks for any body the player has touched. The first body whose
    centre is within $B.r + \text{playerSize}$ of the player becomes
    the new ground body and the player snaps to its surface.
\end{itemize}

\subsection*{Which body's gravity wins}

This is the most interesting new question. The player is now somewhere
in open space, and several bodies might be tugging at them. The game
makes a simple choice: each tick it picks one body --- the dominant
gravity source --- and uses that body's centre to compute the gravity
direction. Two rules pick that body:

\begin{enumerate}
    \item Only bodies whose centre is within a generous influence
    radius of the player (about $4.5$ times their own radius) are
    considered at all. Anything farther exerts no pull.
    \item Among those, the body whose \emph{surface} is closest to
    the player wins.
\end{enumerate}

Using \emph{surface} distance instead of \emph{centre} distance is
important. The big start planet has a much larger surface even though
its centre may be farther from the player than some nearby platform's
centre. Surface distance keeps the player feeling glued to the big
planet while standing on it, and only hands gravity off to the small
platform once the player is genuinely closer to that platform than to
the planet's surface.

\subsection*{Floating platforms}

A mini-platform is just a body with a small radius. Everything that
worked for a planet works for a platform with no extra code: the
player walks around it the same way, gravity points to its centre,
and landing snaps to its surface. This is the payoff of the
``everything is a circular body'' design --- adding a new platform is
one line of data.

\subsection*{Air control}

While in the air, the left/right keys nudge the player's velocity
\emph{tangent} to gravity --- i.e.\ perpendicular to the line from
the player to the current gravity source. The nudge is small and
capped, so the player can steer their arc but can't accelerate
forever. This is enough to make jumps between platforms feel
controllable rather than ballistic.

\subsection*{The launch star}

A launch star is a single point in space with three pieces of state:
its position, a target position, and a ``used'' flag. Touching it and
pressing Space sets the player's velocity to a unit vector aimed at
the target, multiplied by a high speed. The player then flies in a
nearly straight line until gravity from a distant body bends them
into a landing. A short cooldown after firing temporarily disables
landing collisions, so the player doesn't accidentally clip into a
platform that happens to lie on the launch line.

In the world this game ships with, the launch star sits above the
start planet and aims directly at the target planet, so a normal jump
plus Space carries the player across the entire scene in one trip.

\subsection*{The warp tube}

A warp tube has two endpoints and a stored ``exit direction'' for
each. The exit direction is the way the player flies out when they
emerge at that endpoint. Touching either endpoint teleports the
player to the other endpoint, places them just outside it along its
exit direction, and gives them a small velocity in that same
direction. A cooldown prevents the player from re-entering the tube
immediately on exit and bouncing back and forth forever.

In the world, the tube connects the left side of the target planet
back to the right side of the start planet --- a one-step return
journey, so the player can grab stars on the target side and then
warp home rather than relaunching across the gap.

\subsection*{Falling into space}

If the player ever drifts off-screen --- usually by misfiring a jump
or a launch --- they lose a life and respawn on top of the start
planet. This is the simplest possible failsafe and avoids the need
for a kill plane or background drag.

\subsection*{What this adds on top of the single-planet game}

Compared to Mini Galaxy 1, the additions are:

\begin{itemize}
    \item Player position becomes Cartesian, plus a separate ``which
    body am I on'' field for ground state.
    \item A new ``which body's gravity is active'' function based on
    surface distance.
    \item Landing logic that picks up any body the player has
    touched, not just one fixed planet.
    \item Two new world objects --- launch stars and warp tubes ---
    each implemented as a handful of fields and a small block of
    collision code.
\end{itemize}

What remains unchanged is the spine: state lives as plain data, one
big game loop steps it forward each tick, and the screen is redrawn
from that state every frame. Adding a third planet, a new pipe, or a
second launch star would each be a few lines of data --- the engine
underneath does not need to know about them.

\section*{Notes: An Energy Grid Mix Simulator}

This chapter changes register. Instead of another game, this is a
small interactive science tool that uses the same building blocks
(\texttt{pygame}, a single render loop, plain dictionaries for state)
to play with a different question: what happens when you try to
supply a country-sized electricity grid using different mixes of
generating sources? The source file is \texttt{energy\_grid.py}, in
the parent project folder.

\subsection*{What you see}

The window has three parts. At the top, a stacked-area chart spans a
full 24-hour day. Each band is a different source --- yellow for
solar, light blue for wind, dark blue for hydro, red for geothermal,
green for nuclear, olive for biomass, orange for natural gas, brown
for coal. A white line draws the grid's demand on top of the stack.
A vertical white cursor sweeps through the day in real time; one
simulated day takes about 45 seconds.

Below the chart there is one slider per source, setting that
source's installed capacity as a percentage of peak demand. Drag a
slider and the chart updates immediately.

To the right, a stats panel reports totals for the day --- total
demand, total generation, reliability (the fraction of demand that
was actually met), curtailment (the energy generated above demand
that has to be thrown away), renewable share, average CO\textsubscript{2}
intensity, and average cost --- as well as a live ``right now''
readout for the current simulated minute.

\subsection*{The model}

The simulator is a deliberately simple physics-free model. Two
things drive everything that's drawn:

\begin{enumerate}
    \item A \textbf{demand curve} $D(t)$, expressed as a fraction of
    peak demand. It uses a constant base load of about 0.5, plus two
    Gaussian bumps at 08:00 (morning peak) and 19:00 (evening peak)
    --- the standard shape of a residential-plus-commercial load.
    Multiplied by a chosen peak (50\,GW), this gives the demand in
    gigawatts at each moment of the day.

    \item A \textbf{capacity-factor profile} $f_s(t)$ for each
    source $s$, returning the fraction of nameplate capacity that
    that source actually delivers at time $t$. Each source has its
    own characteristic shape, described below.
\end{enumerate}

The instantaneous generation from source $s$ is then just
\[
G_s(t) = C_s \cdot f_s(t),
\]
where $C_s$ is the installed capacity controlled by that source's
slider. Total grid generation at time $t$ is the sum of all $G_s(t)$.

\subsection*{Where each source's shape comes from}

\begin{itemize}
    \item \textbf{Solar PV} is a half-sine peaking at midday, zero
    from sunset to sunrise. The peak amplitude is set so the daily
    average capacity factor lands near 27\%, matching real
    sunny-region utility-scale solar.
    \item \textbf{Wind} is a sum of slowly-varying sines plus a
    small diurnal bias, clamped to a realistic range. It averages
    around 35\%, with bumpy hour-to-hour swings --- not real wind
    data, but the right qualitative behaviour.
    \item \textbf{Hydro} is roughly flat near 45\%, with small bumps
    at the morning and evening demand peaks, representing
    dispatchable peak-shaving.
    \item \textbf{Geothermal} and \textbf{biomass} are flat baseload
    near 78\% and 68\%.
    \item \textbf{Nuclear} is a flat 92\% --- nuclear plants are
    overwhelmingly run at full power because their fuel cost is
    tiny compared to their capital cost.
    \item \textbf{Natural gas} is shaped like a peaker: a base of
    25\% with two Gaussian bumps following the morning and evening
    demand peaks.
    \item \textbf{Coal} is essentially baseload with a slow daily
    swing.
\end{itemize}

\subsection*{Where the numbers in the stats panel come from}

Each source carries two factual numbers used by the stats panel.

\begin{itemize}
    \item \textbf{Lifecycle CO\textsubscript{2} intensity}, in grams
    of CO\textsubscript{2}-equivalent per kWh. These come from IPCC
    AR5 (Working Group III, 2014) and IPCC AR6 (2022) lifecycle
    medians: roughly 11 (wind), 12 (nuclear), 24 (hydro), 38
    (geothermal), 48 (solar PV), 230 (biomass), 490 (natural gas
    combined-cycle), 820 (coal).
    \item \textbf{Levelized cost of electricity}, in USD per MWh.
    Approximate mid-range values from Lazard's LCOE Analysis v16
    (2023) and EIA AEO 2024: $\sim$35 (wind), $\sim$40 (solar PV),
    $\sim$45 (hydro), $\sim$55 (natural gas), $\sim$75
    (geothermal), $\sim$90 (biomass and coal), $\sim$165 (new-build
    nuclear).
\end{itemize}

For each simulated day the panel computes a weighted average,
weighted by energy delivered by each source. So the headline
``g/kWh'' and ``\$/MWh'' figures are the daily averages your chosen
mix would produce.

\subsection*{What the simulator is good for}

A few patterns become obvious within seconds:

\begin{itemize}
    \item Push solar to 150\% and you flood the grid at noon (huge
    curtailment) but leave a yawning gap from sunset to sunrise.
    \item Push wind alone to 100\% and watch the white demand line
    cross repeatedly into the stacked area --- some hours met,
    others underrun, none reliable.
    \item Stack only baseload (nuclear, geothermal, biomass) and
    reliability climbs near 100\% but cost rises and you can't
    follow evening peaks well.
    \item Switch between presets (keys 1, 2, 3) to compare a
    coal-heavy mix, today's average mix, and a renewables-heavy mix
    side by side --- watching CO\textsubscript{2} intensity drop and
    average cost wobble.
\end{itemize}

\subsection*{What the simulator is \emph{not}}

It is important to be honest about how stripped-down this is. The
following are all missing on purpose to keep the model
understandable:

\begin{itemize}
    \item \textbf{No storage}. Batteries, pumped hydro, hydrogen
    --- nothing soaks up midday solar to spend it after dark. In
    practice, storage radically changes how much solar and wind a
    real grid can absorb.
    \item \textbf{No transmission}. The grid is treated as one
    bus. Real grids have geography, congestion, and losses.
    \item \textbf{No reserves or ramp constraints}. Coal plants
    can't actually swing on an hourly basis the way the chart
    implies; gas peakers can but cost more on the margin.
    \item \textbf{No weather or seasonality}. Solar and wind use a
    single archetypal day; real output varies enormously with
    season, latitude, and weather.
    \item \textbf{No interactions on cost}. The displayed LCOEs are
    flat averages, but the marginal cost of dispatching one more MWh
    of a given source depends on the rest of the mix.
\end{itemize}

So the headline numbers should be read as ``roughly the right order
of magnitude, with the right relative ranking,'' rather than as
predictions.

\subsection*{How it shares DNA with the games}

Underneath the chart and sliders, the loop is identical to the games
in earlier chapters:

\begin{enumerate}
    \item Read inputs (mouse drags on sliders; number keys to apply
    presets; Space to pause the clock).
    \item Update state (advance the simulated clock; recompute
    generation from the current slider values).
    \item Compute derived stats (totals, reliability, curtailment,
    averages).
    \item Draw everything from scratch.
\end{enumerate}

The state is just a list of source dictionaries with a
\texttt{capacity\_pct} field and a callable \texttt{profile}. There
is no scene graph, no widget framework, and no plotting library ---
the chart is hand-drawn as a stack of polygons over a 24-hour axis,
in the same way the snake game drew its squares. The point is that
the same minimal toolkit that runs little games is enough to build
a small interactive science visualisation.

\section*{Notes: Energy Storage Technologies}

A natural follow-on to the grid mix simulator is the question of
\emph{storage}: what soaks up the surplus when the chart goes above
the demand line, and what fills the gap when it goes below? This
chapter inventories the main grid-scale storage technologies, lists
the numbers that are reasonably well-established in the open
literature, walks through how the efficiency-over-time figure was
computed, and lists what I had to flag as unknown.

The interactive tool that goes with this chapter is
\texttt{storage\_compare.py}; it plots round-trip efficiency vs
storage duration on a log-time axis, with togglable curves per
technology.

\subsection*{What I am and am not modelling}

This chapter and the tool both isolate one effect:
\textbf{self-discharge during idle storage}. The model is

\[
\eta(t) \;=\; \mathrm{RTE} \cdot (1 - r_{\mathrm{sd}})^{t / (1\ \text{day})}
\]

where:

\begin{itemize}
    \item $\mathrm{RTE}$ is the round-trip electrical efficiency at
    $t=0$, combining charging and discharging losses.
    \item $r_{\mathrm{sd}}$ is the equivalent exponential
    self-discharge rate per day.
\end{itemize}

What this model leaves out, on purpose:

\begin{itemize}
    \item \textbf{Calendar degradation} --- the slow permanent loss
    of capacity in a cell over years (typically $\sim$1--3 \%/yr for
    Li-ion), even when idle.
    \item \textbf{Cycle degradation} --- permanent capacity loss per
    full charge/discharge cycle.
    \item \textbf{Operational complications} --- ramp rates,
    response times, partial-state-of-charge effects, thermal
    management overhead, etc.
    \item \textbf{Economic crossover} --- in practice the choice
    between storage technologies at long durations is dominated by
    \emph{capital cost per kWh of energy capacity}, not by
    efficiency. A hydrogen salt cavern stores cheaply at scale even
    though its round-trip efficiency is poor.
\end{itemize}

So $\eta(t)$ answers ``of the kilowatt-hour I put in today, how many
do I get back out $t$ later?'' If you also need the answer to ``and
how much did the storage system cost in capital and consumables to
deliver that kWh?'', see the LCOS column of the table below.

\subsection*{The data}

The table below gives the values used by the tool, with their ranges
where literature spread is significant. Confidence indicates whether
the value comes from commercial deployments (high), demonstrators
(moderate) or vendor claims only (low). All numbers are mid-range
estimates from cited public reports; verify against the originals
for application-critical use.

\begin{footnotesize}
\begin{tabular}{|l|c|c|l|l|l|l|}
\hline
\textbf{Technology} & \textbf{RTE} & \textbf{Self-d.}
  & \textbf{Duration} & \textbf{Cost \$/kWh} & \textbf{Lifetime}
  & \textbf{Confid.} \\
\hline
Li-ion (LFP)        & 90--95 \% & 2--3 \%/mo  & 1--8 h        & 150--300   & 10--15 yr / 3k--10k cyc & high \\
Lead-acid           & 75--85 \% & 3--5 \%/mo  & 1--6 h        & 100--250   & 5--15 yr / 500--1.5k cyc & high \\
Vanadium flow       & 65--75 \% & $<$2 \%/mo  & 4--12 h       & 300--700   & 20+ yr / 15k+ cyc      & high \\
Pumped hydro (PHS)  & 70--85 \% & $\sim$0     & 4--24+ h      & 50--200    & 50--100+ yr            & high \\
Diabatic CAES (gas) & 42--55 \% & $<$1 \%/mo  & 4--24 h       & 50--150    & 25+ yr                 & high\textsuperscript{a} \\
Adiabatic CAES      & 60--70 \% & $\sim$5 \%/d & hours         & TBD        & 25+ yr (est.)          & low \\
Flywheel            & 85--95 \% & 10 \%/hr   & sec--min      & 1k--5k     & 20+ yr / 100k+ cyc     & high \\
Molten salt (P$\to$P)& 30--45 \%& 1--2 \%/d   & 4--12 h       & 30--80     & 30+ yr                 & high\textsuperscript{b} \\
Supercapacitor      & 95--98 \% & 5--40 \%/d  & sec--hr       & 1k--10k    & 20+ yr / 1M+ cyc       & high \\
Hydrogen (P$\to$H$_2\to$P) & 30--45 \% & $<$0.1 \%/d & d--mo & 5--30 (gas)/varies & varies & moderate \\
Liquid air (LAES)   & 50--70 \% & $\sim$1 \%/d & h--d          & 200--500   & 25+ yr (est.)          & low \\
Gravity (solid)     & 80--90 \% & $\sim$0     & h             & TBD        & TBD                    & low \\
\hline
\end{tabular}

\medskip\noindent
\textsuperscript{a}Two commercial plants operating since 1978 (Huntorf,
Germany) and 1991 (McIntosh, USA). \\
\textsuperscript{b}Demonstrated extensively as a thermal store in
concentrated-solar-power plants. Round-trip to electricity requires
pairing with a heat engine, which is where the efficiency loss sits.
\end{footnotesize}

\subsection*{Sources}

The numeric ranges above are drawn from publicly available reports.
The most directly useful, with the metrics they typically supply:

\begin{itemize}
    \item \textbf{IEA Energy Technology Perspectives} (latest 2024) ---
    cross-technology storage cost and performance.
    \item \textbf{IEA Batteries and Secure Energy Transitions} (2024)
    --- battery RTE, cost trajectories.
    \item \textbf{IRENA Electricity Storage and Renewables: Costs and
    Markets to 2030} (2017) --- foundational cross-technology
    reference.
    \item \textbf{IRENA Innovation Outlook: Thermal Energy Storage}
    (2020).
    \item \textbf{NREL Annual Technology Baseline} (yearly; 2024
    edition) --- utility-scale storage parameters.
    \item \textbf{NREL Cost Projections for Utility-Scale Battery
    Storage: 2023 Update} (Cole \& Karmakar).
    \item \textbf{Lazard's Levelized Cost of Storage Analysis} v8
    (2023) --- LCOS by duration and technology.
    \item \textbf{US DOE Energy Storage Grand Challenge: Storage Market
    Report 2022}.
    \item \textbf{IEC White Paper: Electrical Energy Storage} (2011) ---
    older, but still useful technology taxonomy.
    \item \textbf{Sandia / DOE OE Global Energy Storage Database} ---
    operational fleet data, useful for sanity-checking ranges.
\end{itemize}

I have \emph{not} retrieved these documents fresh while writing this
chapter; the numbers are values I recognise from reading these
reports historically. Treat the table as an at-a-glance summary
suitable for orientation, and re-pull the exact figures from the
above sources before relying on them.

\subsection*{Where values are missing}

Three rows have ``TBD'' or ``low confidence'' entries; these reflect
genuine gaps in the public record at scale, not omissions:

\begin{itemize}
    \item \textbf{Adiabatic CAES} (e.g.\ ADELE, RICAS2020): no
    full-scale commercial plant exists. Stated round-trip figures are
    from demonstrators and design studies. Heat-store self-discharge
    is the dominant practical loss and depends entirely on insulation
    quality.
    \item \textbf{Gravity storage} (e.g.\ Energy Vault, Gravitricity):
    a small number of demonstrators have been built; published RTE
    figures are vendor claims and have not, as far as I know, been
    independently audited at grid scale.
    \item \textbf{LAES}: only one commercial plant (Highview Power's
    50 MW / 250 MWh CRYOBattery near Manchester, UK) is coming online
    in 2024--2025. Lifetime and long-term cost data therefore do not
    exist yet.
\end{itemize}

If the simulator shows a curve for one of these, treat it as
``best public estimate'' rather than ``measured.''

\subsection*{How the efficiency curves are computed}

The curve for each technology is just the formula above, evaluated
at $\sim$240 logarithmically-spaced time points between 1 minute and
10 years. Two examples worked out:

\paragraph{Lithium-ion (LFP) after one month.}
$\mathrm{RTE} = 0.92$, $r_{\mathrm{sd}} = 0.025 / 30 \approx 8.3 \times 10^{-4}$
per day. After 30 days,
\[
\eta(30\ \text{d}) = 0.92 \cdot (1 - 8.3\times 10^{-4})^{30}
 = 0.92 \cdot 0.975 = 0.897.
\]
So you get back about 89.7 \% of what you put in after one month --- the
round-trip dominates and the self-discharge is barely 2 \%.

\paragraph{Lithium-ion (LFP) after one year.}
\[
\eta(365\ \text{d}) = 0.92 \cdot (1 - 8.3\times 10^{-4})^{365}
 = 0.92 \cdot 0.739 = 0.680.
\]
At a year of idle storage, self-discharge has now eaten about a
quarter of the stored energy, and only 68 \% of the input
kilowatt-hour comes out.

\paragraph{Flywheel after one day.}
$\mathrm{RTE} = 0.88$, $r_{\mathrm{sd}} = 1 - (1 - 0.10)^{24} \approx 0.920$
per day (10 \%/hour). After one day,
\[
\eta(1\ \text{d}) = 0.88 \cdot (1 - 0.920) = 0.88 \cdot 0.080 = 0.070.
\]
A flywheel left to idle for 24 hours returns 7 \% of what was put in:
this is why flywheels are not used for daily storage even though their
round-trip is excellent.

\subsection*{Crossover points worth knowing}

\paragraph{Lithium-ion vs pumped hydro.}
PHS has a lower RTE (78 \% vs 92 \%) but essentially no
self-discharge. Setting their efficiencies equal:
\[
0.92 \cdot (1 - 8.3\times 10^{-4})^d = 0.78,
\]
\[
(0.99917)^d = 0.848, \quad d = \frac{\ln 0.848}{\ln 0.99917} \approx 199\ \text{days}.
\]
So Li-ion is more efficient than pumped hydro for storage durations
\emph{below} about 6.5 months; PHS wins for longer storage. (In
practice this comparison is rarely the binding one --- PHS sites are
geographically constrained and Li-ion is rarely sized for
month-long discharge.)

\paragraph{Lithium-ion vs hydrogen.}
$\mathrm{RTE}_{\mathrm{H}_2} = 0.35$, $r_{\mathrm{sd}} \approx 0.0005$/day.
Setting equal:
\[
0.92 \cdot (0.99917)^d = 0.35 \cdot (0.9995)^d
\]
\[
\left(\frac{0.99917}{0.9995}\right)^d = \frac{0.35}{0.92}
\]
\[
(0.99967)^d = 0.380, \quad d = \frac{\ln 0.380}{\ln 0.99967} \approx 2929\ \text{days} \approx 8\ \text{years}.
\]
On efficiency alone, hydrogen never beats lithium for storage
durations under about eight years. Yet hydrogen is the leading
candidate for \emph{seasonal} energy storage, because its
\textbf{capital cost per stored kWh} is one to two orders of
magnitude lower than batteries at scale. The point is that for long
durations, the right question is dollars per stored kWh, not
percentage points of efficiency.

\subsection*{Headline risks per technology}

\begin{itemize}
    \item \textbf{Li-ion}: thermal runaway and fire. NMC chemistries
    are more prone than LFP; large-format installations have suffered
    headline incidents (APS McMicken, AZ 2019; Beijing 2021;
    Moss Landing, CA 2025). Fire mitigation engineering is now
    standard.
    \item \textbf{Lead-acid}: lead toxicity in manufacture and
    end-of-life; sulfation when left discharged.
    \item \textbf{Vanadium flow}: electrolyte handling; price
    volatility of vanadium pentoxide.
    \item \textbf{Pumped hydro}: geographic and ecological
    constraints; long permitting timelines; dam-failure tail risk
    (very small but high-consequence).
    \item \textbf{Diabatic CAES}: burns natural gas to reheat
    expanding air, so its full-chain greenhouse emissions are non-zero
    even though its electricity input is clean.
    \item \textbf{Flywheel}: catastrophic mechanical failure if the
    rotor fails at speed (containment vessels are designed for this,
    but the failure mode is real).
    \item \textbf{Molten salt}: corrosion and freezing of the salt
    medium; the heat engine paired with it carries its own risks.
    \item \textbf{Supercapacitor}: very low energy density makes it
    a poor primary store; usually paired with batteries.
    \item \textbf{Hydrogen}: flammability and detonability;
    embrittlement of containment; small molecules leak through
    standard fittings; emerging research suggests leaked H$_2$ itself
    has an indirect global-warming effect by extending methane's
    lifetime.
    \item \textbf{LAES}: nascent technology; long-term reliability
    unknown.
    \item \textbf{Gravity storage}: limited deployment evidence;
    economics at grid scale unverified.
\end{itemize}

\subsection*{Reading the chart}

In the tool, drag the white cursor along the time axis to see each
technology's efficiency at any duration. Three regimes become
visually obvious:

\begin{itemize}
    \item \textbf{Seconds to minutes:} supercapacitors and flywheels
    are at their best; chemical batteries are also fine; thermal,
    PHS and CAES are over-engineered for these timescales.
    \item \textbf{Hours to one day:} Li-ion and flow batteries, PHS
    and CAES, and thermal all sit in the same broad band. This is
    where the most competitive market exists today.
    \item \textbf{Weeks to months:} only PHS and hydrogen remain
    competitive. Batteries' self-discharge eats them; flywheels and
    supercapacitors are long gone; thermal has cooled off.
\end{itemize}

The crossover analysis above tells you exactly when in days each
pair of curves meets. The economic case usually swings sooner than
the efficiency case --- but if your application really is ``charge
once, sit for months, discharge once'' (e.g.\ seasonal solar in a
high-latitude country), efficiency does matter, and the chart
shows why the long-duration question keeps coming back to chemical
fuels and pumped hydro.

\subsection*{Revision: efficiency values cross-checked against real installations}

The first version of this chapter used cell-level or DC efficiencies
for some technologies and omitted the parasitic loads that real
grid-connected systems consume continuously. That overstated several
technologies, most importantly lithium-ion at multi-month durations.
The values used in the tool have been revised to be \emph{AC-to-AC,
system-level, including parasitic loads}, anchored to specific
real-world implementations. The changes:

\begin{footnotesize}
\begin{tabular}{|l|c|c|l|}
\hline
\textbf{Technology} & \textbf{RTE (was)} & \textbf{RTE (now)}
  & \textbf{Anchor / reason} \\
\hline
Li-ion (LFP)        & 92 \% & 86 \% & NREL ATB 2024 / Lazard LCOS v9 utility-scale, AC-AC \\
Lead-acid           & 80 \% & 78 \% & AC-AC system level; older spec sheets are DC \\
Vanadium flow       & 70 \% & 68 \% & Pump parasitic during operation (3--5 \%) \\
Pumped hydro        & 78 \% & 78 \% & Unchanged. Bath County $\sim$72 \%, Goldisthal $\sim$80 \% \\
CAES diabatic       & 52 \% & 48 \% & Huntorf $\sim$42 \%, McIntosh $\sim$54 \% $\to$ mid $\sim$48 \% \\
Adiabatic CAES      & 65 \% & 65 \% & Unchanged. No commercial plant; design estimate \\
Flywheel            & 88 \% & 86 \% & Beacon Power / Amber Kinetics fleet \\
Molten salt P$\to$P & 40 \% & 40 \% & Unchanged. Note: no commercial pure P$\to$P plant exists \\
Supercapacitor      & 96 \% & 93 \% & DC cell $\sim$96 \%, AC-AC drops with inverters \\
Hydrogen            & 35 \% & 35 \% & Unchanged. PEM 70 \% $\times$ compr.\ 90 \% $\times$ FC 55 \% \\
LAES                & 60 \% & 52 \% & Highview pilot demonstrated $\sim$50 \%; 60--70 \% is design w/ waste heat \\
Gravity (solid)     & 80 \% & 75 \% & Vendor claims 75--85 \%; no independently audited grid plant \\
\hline
\end{tabular}
\end{footnotesize}

The self-discharge values for battery technologies were also raised
to include the BMS, cooling and (for flow batteries) pump parasitic
loads that run continuously while the system sits idle:

\begin{itemize}
    \item Li-ion: $\sim$2.5 \%/mo (cell) $\to$ $\sim$4 \%/mo
    (cell + BMS + thermal management).
    \item Lead-acid: $\sim$4 \%/mo $\to$ $\sim$6 \%/mo
    (cell + float charge maintenance).
    \item Vanadium flow: $\sim$1 \%/mo $\to$ $\sim$3 \%/mo
    (near-zero chemical + continuous pump parasitic).
\end{itemize}

\paragraph{Specific real-world anchors.}
The most concrete data points behind the revised numbers:

\begin{itemize}
    \item \textbf{Huntorf CAES} (Germany, 290 MW, since 1978):
    measured RTE around 42 \% when fairly accounting for the natural
    gas combusted during expansion.
    \item \textbf{McIntosh CAES} (Alabama, 110 MW, since 1991):
    around 54 \% thanks to its recuperator. Average of the two
    commercial CAES plants ever built sits near 48 \%.
    \item \textbf{Bath County PHS} (Virginia, 3.0 GW, 1985): the
    largest pumped-hydro plant in the world; reported round-trip
    around 70--75 \%. Modern variable-speed PHS plants such as
    \textbf{Goldisthal} (Germany, 2003) and \textbf{Linthal}
    (Switzerland, 2017) reach 80 \%+. The fleet mid-range is therefore
    $\sim$78 \%.
    \item \textbf{Beacon Power Stephentown} (NY, 20 MW flywheel):
    AC-AC RTE around 85 \% for fast frequency-regulation service.
    \item \textbf{Highview Power Pilsworth/Bury} (UK, 5 MW LAES
    pilot): demonstrated RTE in the 50 \% range. The follow-on
    50 MW Carrington plant is designed for 60--70 \% \emph{but only
    when waste heat from a neighbouring industrial site is
    integrated}; standalone the design figure is closer to 50--55 \%.
    \item \textbf{Energy Vault EVx Rudong} (China, 25 MW / 100 MWh
    gravity, online 2024): vendor-stated round-trip in the 75--80 \%
    range. No independent audit available yet at the time of writing.
    \item \textbf{NREL Annual Technology Baseline 2024}: utility-scale
    4-hour Li-ion at 85 \% AC-AC RTE.
    \item \textbf{Lazard's LCOS v9 (2024)}: uses around 86 \% AC-AC
    for utility-scale LFP, consistent with NREL.
\end{itemize}

\paragraph{What I am still uncertain about.}
A few values remain noisier than the table suggests:

\begin{itemize}
    \item \textbf{Adiabatic CAES} is fundamentally an estimate.
    Hydrostor's first commercial-scale plants (Australia, Canada)
    are under construction but not yet operating, so the 65 \% figure
    is from design studies, not from operating fleet.
    \item \textbf{LAES} has exactly one commercial plant coming
    online; the demonstrated 50 \% and designed 60--70 \% bracket the
    real number, and where it lands depends on whether waste-heat
    integration is available.
    \item \textbf{Gravity storage} only has one nameplate-scale plant
    operating, and only the vendor has published efficiency data
    on it. Treat the 75 \% as ``vendor-claimed pending audit.''
\end{itemize}

\paragraph{What the revised chart now shows.}
With Li-ion dropped from 92 \% to 86 \% and its self-discharge raised
to include BMS parasitic, the multi-month region of the chart is
much more honest. Pumped hydro now surpasses Li-ion for storage
durations beyond roughly 60--70 days (versus the old $\sim$200 days),
which matches operational reality: nobody actually uses Li-ion for
month-scale storage, and the chart now shows why.
