\documentclass[conference]{IEEEtran}
\usepackage[utf8]{inputenc}
\usepackage{cite}
\usepackage{amsmath}
\usepackage{hyperref}

\title{Power System Stability in a Decarbonising Europe:\\
From the Inertia Problem to AI-Enabled Operation}

\author{\IEEEauthorblockN{Judith Drack}\\
\IEEEauthorblockA{Draft prepared with assistance from an AI writing tool}}

\begin{document}
\maketitle

\begin{abstract}
Decarbonising the electricity sector requires replacing synchronous generators with inverter-based renewable resources. This shift undermines a structural property of conventional grids---the rotational inertia stored in the rotating masses of synchronous machines---and replaces an electromechanically-defined system with one whose behaviour is set by power-electronic controls. The European synchronous area is among the most advanced examples of this transition. This paper explains, first, the physical inertia problem and its operational consequences; second, what it means for Europe's grid to evolve from a synchronous-machine-dominated to a power-electronics-dominated system; and third, how artificial intelligence (AI) and machine learning (ML) can support stability during decarbonisation through forecasting, state estimation, dynamic security assessment, real-time control, and predictive maintenance. The limitations and governance challenges of applying AI in safety-critical grid operation are discussed.
\end{abstract}

\section{Introduction}\label{sec:intro}
The European Union has committed in law to net-zero greenhouse-gas emissions by 2050 under the European Climate Law~\cite{eu2021climatelaw}. Electricity is the central enabler of this commitment, both because it is the energy carrier that can most plausibly be decarbonised at scale and because the electrification of transport, heating and industry will shift large new loads onto the grid. The expansion of renewable generation needed to meet these targets must be matched by an equally rapid transformation of how the grid is operated: every increment of variable, inverter-based generation displaces a piece of the physical infrastructure that has historically kept the system stable. Decarbonisation is, in this sense, also a stability problem---the binding constraint on how quickly thermal generation can be retired is not the size of the renewable build-out but the operational margins of the network that connects it to demand.

This paper is structured around three questions. Section~\ref{sec:stability} reviews what power-system stability means in the modern framework. Section~\ref{sec:inertia} explains the inertia problem that arises when synchronous machines are displaced. Section~\ref{sec:transition} examines what it means for the European grid to evolve from a system dominated by synchronous machines to one dominated by power electronics. Section~\ref{sec:ai} surveys how AI and ML are being applied to maintain and improve stability during this transition, with attention to the decarbonisation end-goal. Section~\ref{sec:risks} discusses limitations, risks and governance, and Section~\ref{sec:conclusion} concludes.

\section{Power System Stability}\label{sec:stability}
The classical definition of power-system stability, codified by the IEEE/CIGRE joint task force on stability terms, identifies three categories: rotor-angle, frequency, and voltage stability, each subdivided by time scale (short, medium, long term) and disturbance magnitude (small, large)~\cite{kundur1994,kundur2004definition}. \emph{Rotor-angle} stability concerns the ability of synchronous machines to remain in synchronism after a disturbance and is governed by the swing equation that relates net electrical torque to rotor acceleration. \emph{Frequency} stability concerns the ability of the system to keep frequency within an acceptable band following a generation-load imbalance, and is supported by the inertial response of rotating masses and by primary, secondary, and tertiary control. \emph{Voltage} stability concerns the ability to maintain acceptable voltages at all buses, and depends critically on the supply of reactive power.

In 2021 the IEEE/CIGRE task force extended this taxonomy with two new categories that reflect the increasingly electronic character of the grid: \emph{converter-driven} stability, addressing fast and slow interactions of inverter control loops with the network, and \emph{resonance} stability, addressing sub-synchronous oscillations involving converter controls, series compensation, and turbine shafts~\cite{hatziargyriou2021revisited}. The conceptual shift is significant: in a synchronous-machine-dominated grid, stability is largely a property of physics, whereas in a converter-dominated grid it becomes a property of \emph{control}.

\section{The Inertia Problem}\label{sec:inertia}
Each synchronous generator stores kinetic energy in its rotor,
\begin{equation}
E_{\mathrm{kin}} = \tfrac{1}{2} J \omega^{2},
\end{equation}
where $J$ is the moment of inertia and $\omega$ the rotational angular velocity. Because the rotor is electromagnetically coupled to the grid, any imbalance between mechanical input and electrical output power immediately accelerates or decelerates the rotor and so changes the system frequency. The dynamics are captured by the \emph{swing equation},
\begin{equation}
2H\,\frac{df}{dt} = P_{\mathrm{mech}} - P_{\mathrm{elec}},
\end{equation}
where $H$ is the inertia constant (energy stored per unit MVA) and $f$ the frequency. The instantaneous response is automatic: no controller needs to act for inertia to slow a frequency excursion. This automatic response buys the time required for primary frequency control (governor response) to engage, typically within a few seconds, and thereby keeps the rate of change of frequency (RoCoF) within bounds for which protection schemes have been designed~\cite{kundur1994}.

Inverter-based resources such as photovoltaic plants, modern wind turbines, and battery storage are interfaced to the grid through power-electronic converters. Their output is set by control loops rather than by rotor dynamics, and absent specific provisions they make no inertial contribution. As IBRs displace synchronous machines, the system-wide inertia constant falls, RoCoF following a contingency rises, and the time available for controllers, protection, and load-frequency-control schemes to act shrinks~\cite{tielens2016relevance,ulbig2014impact,milano2018foundations}. ENTSO-E has explicitly recognised this trajectory in its operational studies and has updated its RoCoF tolerances accordingly~\cite{entsoe2020inertia}.

The operational consequences extend beyond frequency response. Synchronous machines also supply the short-circuit current that protection relays use to detect and clear faults; the loss of short-circuit current as the machine fleet retires complicates protection coordination. Under-frequency load-shedding schemes calibrated for historic inertia levels may act too late or not at all under faster RoCoF, raising the probability of cascading failures. Extreme weather events, increasing in frequency and severity, additionally stress the system through simultaneous demand spikes and equipment outages, as documented in the joint FERC/NERC inquiry into the February 2021 Texas cold-weather outages~\cite{fercnerc2021texas}. NREL has framed inertia not as an end in itself but as one of several services---inertial response, primary frequency response, ramping capability, voltage support---that must continue to be provisioned as the resource mix changes~\cite{denholm2020inertia,nerc2015erstf}.

\section{From Synchronous-Machine-Dominated to Power-Electronics-Dominated Grids}\label{sec:transition}
The Continental European synchronous area, operated jointly by ENTSO-E members, is among the world's largest interconnected grids. Historically, its stability has been underwritten by a dense fleet of large thermal and nuclear synchronous generators, complemented by hydro. The decarbonisation pathway codified by the European Climate Law~\cite{eu2021climatelaw} replaces a substantial share of these machines with wind, photovoltaic, and storage assets that interface to the grid through power electronics. The question of how to operate a synchronous area in which converters, rather than rotating machines, are the dominant source of generation is no longer hypothetical.

Several physical and operational properties change simultaneously in this transition.

\subsection{Loss of Inertia}
As discussed in Section~\ref{sec:inertia}, the time scale on which the system can absorb imbalances passively shortens. The system inertia constant is no longer a stable, plant-design property; it varies hour to hour with the dispatch, and even from one frequency event to the next.

\subsection{Falling Short-Circuit Current}
Synchronous machines deliver large fault currents that protection relays rely on to identify and isolate faults. Converters limit their fault current to protect their semiconductors, typically to 1.1--1.5\,pu of nominal rating. As the IBR share rises, short-circuit current at fault locations falls, which reduces the visibility of faults to protection relays and may render distance-protection settings unreliable~\cite{entsoe2017hpopeips}.

\subsection{Control-Defined Dynamics}
Synchronous-machine grids respond to disturbances through the slow electromechanical dynamics of governors and exciters; converter-dominated grids respond through the much faster bandwidth of inverter control loops. The dynamics on millisecond time scales now matter for stability, exposing new failure modes: converter-driven instabilities and sub-synchronous interactions between converters and the network impedance, formally recognised in the 2021 stability taxonomy~\cite{hatziargyriou2021revisited}.

\subsection{Grid-Following versus Grid-Forming Behaviour}
Conventional IBRs operate in \emph{grid-following} mode, synchronising to a measured voltage waveform that the synchronous machines establish. As the share of synchronous machines falls, the assumption that there is always a strong reference voltage to follow breaks down. \emph{Grid-forming} inverter control, in which the inverter synthesises an internal voltage reference and can support a network without other generators, is the principal hardware-level response to this loss of system strength~\cite{lasseter2020gridforming}.

\subsection{System-Level Implications}
ENTSO-E has framed the high penetration of power-electronic-interfaced power sources as a system-wide design challenge requiring updates to network codes, planning procedures, and operational practices~\cite{entsoe2017hpopeips}. The implication is that the European grid cannot be operated in the future the way it was operated in the past: stability must be engineered into the control of inverters and orchestrated across distributed resources, rather than inherited from the physics of the machine fleet. The operational complexity of this regime---faster dynamics, more devices, more contingencies, more dispatch options---is beyond what classical, hand-tuned methods were designed for, which is what motivates the application of AI and ML to grid operation.

\section{AI for Grid Stability in a Decarbonising System}\label{sec:ai}
Recent surveys identify several domains in which AI and ML are being applied to support stability under decarbonisation~\cite{duchesne2020ml,ahmad2021ai}. The unifying theme is that learned models can extract more useful structure from operational data than classical methods can, and can deliver answers fast enough to be used inside real-time control loops.

\subsection{Forecasting of Generation, Demand and Flexibility}
Accurate short-term forecasts of renewable output, demand, and flexible-resource availability are the most direct way to reduce uncertainty in operational planning. Deep-learning architectures---recurrent, convolutional and graph-based---outperform classical statistical methods on intra-day and day-ahead horizons, particularly for wind and photovoltaic generation~\cite{ahmad2021ai}. Better forecasts reduce the operating reserve required to cover renewable variability, which in turn allows operators to commit less synchronous backup while still meeting reliability standards. This is one of the few mechanisms by which ML directly enables \emph{deeper} renewable penetration rather than merely managing its consequences. Reinforcement-learning agents have also been used to dispatch demand response and behind-the-meter storage so as to smooth net load~\cite{vazquez2019rl}, turning the explosion of distributed flexible loads from a stability liability into a stability resource.

\subsection{State Estimation and Anomaly Detection}
European transmission networks are increasingly instrumented with phasor measurement units (PMUs) producing high-rate synchronised samples that classical state estimators were not designed to exploit fully. ML models can fuse PMU streams with traditional SCADA measurements to estimate system state, identify bad data, and flag cyber-physical anomalies in time scales suitable for real-time use~\cite{duchesne2020ml}. Graph neural networks have been used for fast contingency screening, identifying which lines or nodes are most exposed under a given dispatch and allowing operators to prioritise mitigations~\cite{donnot2018contingencies}. The value of such methods is especially high in a low-inertia regime, where the operator has less time to react and therefore needs more precise advance warning.

\subsection{Dynamic Security Assessment and Inertia Estimation}
Classical dynamic security assessment (DSA) relies on offline simulations of contingencies and is too slow to track the rapidly changing dispatch of a renewable-heavy grid. ML surrogates can approximate the results of full simulations sufficiently well to be used for online DSA, with explicit attention to interpretability so that the outputs are auditable by operators and regulators~\cite{cremer2019interpretable}. ML methods are also being used to \emph{estimate} system inertia in near-real-time from frequency-event measurements: since the inertia constant is now a time-varying quantity that depends on which converters are connected and in which mode they are operating, estimating it dynamically is itself an operational requirement~\cite{tielens2016relevance}.

\subsection{Real-Time and Topology Control}
Optimal-power-flow and security-constrained dispatch problems become combinatorially harder as the share of converters increases and as operators are asked to consider more contingencies and more topologies. Reinforcement learning is being investigated as a tool for high-dimensional, time-critical control where classical formulations are too slow~\cite{glavic2017rl}. The L2RPN (``Learning to Run a Power Network'') challenges, organised by the French transmission system operator RTE, have shown that RL agents can learn topology-control policies that keep simulated grids within operational limits under contingencies, indicating a path toward learned assistants for control-room operators~\cite{marot2020l2rpn}.

\subsection{Smart Inverter Controls}
The grid-forming control paradigm is the principal hardware-level response to declining system strength~\cite{lasseter2020gridforming}, but it places significant tuning and coordination demands on its controllers. ML is being integrated into these controllers to adapt their parameters to changing grid conditions---network impedance, short-circuit ratio, and the behaviour of neighbouring converters---and to coordinate large populations of converters in a way that classical, decentralised control struggles to do.

\subsection{Predictive Maintenance and Asset Management}
Equipment failure on transmission and distribution networks is a common precipitating cause of cascading instability. ML applied to vibration sensors, dissolved-gas analysis, partial-discharge waveforms and weather data can predict transformer, switchgear and line failures days to weeks ahead, reducing both unplanned outages and the conservative maintenance margins that previously absorbed this risk~\cite{ahmad2021ai}. As the European asset base ages and as climate change increases weather stress on equipment, this margin of foresight becomes a meaningful contributor to stability.

\section{Limitations, Risks, and Governance}\label{sec:risks}
The case for AI in grid operation comes with substantial caveats. ML models are only as good as their training distribution, and renewable assets, load patterns and topology in the European grid change faster than typical training data are refreshed---raising the risk of \emph{distribution shift} in safety-critical operation. The most accurate ML methods are also difficult to interpret, which conflicts with regulatory and operational expectations of explainable, auditable decisions; work on interpretable surrogate models is one direct response~\cite{cremer2019interpretable}.

ML pipelines also introduce new cyber-physical attack surfaces, both through the data feeds they depend on and through the increased connectivity of distributed assets. The European NIS\,2 Directive raises the cybersecurity baseline for essential services, including electricity~\cite{eu2022nis2}. The EU AI Act classifies AI systems used in the management and operation of critical infrastructure as ``high-risk'' and imposes corresponding requirements for risk management, data governance, transparency, human oversight, and accuracy and robustness~\cite{eu2024aiact}; AI for grid operation will therefore have to be deployed within a more demanding regulatory frame than typical commercial ML. The International Energy Agency has more broadly emphasised that the energy and digital sectors must co-evolve, and that AI deployment in critical infrastructure requires deliberate governance rather than incidental adoption~\cite{iea2023ai}.

A practical implication is that the most credible near-term applications of AI in grid operation will be \emph{hybrid}: ML models embedded in or constrained by physics-based formulations, used as accelerators or surrogates rather than as drop-in replacements for the existing control stack. This pattern aligns with regulatory emphasis on explainability and with operators' need to retain understanding of failure modes.

\section{Conclusion}\label{sec:conclusion}
Decarbonising the European electricity supply is, at its core, a problem of operating a power system whose physical character is changing fundamentally. Synchronous-machine-dominated grids derive their stability from the rotational inertia, fault-current capability and electromechanical dynamics of their generators; power-electronics-dominated grids must derive equivalent stability from control rather than physics. The inertia problem is the most widely discussed of the resulting challenges, but it is symptomatic of a broader shift in which classical, hand-tuned operational methods are being asked to manage a system whose complexity exceeds their design assumptions.

AI and ML complement, rather than replace, the physical mitigations being deployed: grid-forming inverters, synchronous condensers, fast storage, and stronger interconnection. Their contribution lies in extracting more reliability from the existing system---through better forecasting, faster state estimation, smarter dispatch, learned topology control, and earlier fault detection---while operating within the safety-critical and regulatory frame that grid operation requires. Whether AI delivers on this promise in practice will depend less on algorithmic progress than on data quality, robust integration with physics-based methods, governance under instruments such as the EU AI Act, and the cumulative judgement of the operators who remain accountable for the result.

\bibliographystyle{IEEEtran}
\bibliography{judith}

\end{document}
